---
title: "Demographic Variation in U.S. Household Income, 2023-2024"
subtitle: "Stat 184 Course Project"
author: "William Dong, Bingjie Yan, Ethan Tluszcz"
date: "May 6, 2026"
date-modified: now
format:
  pdf:
    toc: false
    number-sections: true
    number-depth: 5
    fig-align: center
    cap-location: top
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
    colorlinks: true
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
          commandchars=\\\{\},
          breaklines,
          breaknonspaceingroup,
          breakanywhere
        }
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
csl: apa7.csl
---

```{r}
#| label: setupPackages
#| include: false

# ---- Load Required Packages ----
# Style Guide: BOAST/Tidyverse (https://style.tidyverse.org/)
# Coding standards: snake_case for column names, camelCase for variable
# names, native pipe |>

library(tidyverse)
library(tinytex)
library(knitr)
library(kableExtra)
library(scales)
```

# Introduction

Household income is one of the most widely used measures of economic
well-being in the United States, but the headline national figures
mask substantial variation across demographic groups. The U.S. Census
Bureau's Current Population Survey Annual Social and Economic
Supplement (CPS ASEC) publishes detailed income summaries that allow
researchers to compare how median household income varies by
educational attainment, gender, age of householder, and other
characteristics [@censusBureau2025].

This report conducts an exploratory data analysis on U.S. household
income for 2023 and 2024, focusing on three characteristics:
educational attainment, gender of householder, and age of
householder. Our goal is to identify which characteristics are
associated with the largest disparities in median household income
and which groups experienced meaningful year-over-year change.

# Data and Methods

## Data Source and Provenance

The data come from the Census Bureau report *Income in the United
States: 2024* (Report Number P60-285), Table A-1: *Income Summary
Measures by Selected Characteristics, 2023 and 2024*. Income figures
are reported in 2024 dollars, adjusted using the Chained Consumer
Price Index for All Urban Consumers (C-CPI-U), and households are
surveyed as of March of the following year. The underlying data are
collected by the Census Bureau and the Bureau of Labor Statistics
through the Annual Social and Economic Supplement to the Current
Population Survey, a long-running household survey that has been the
primary source of official U.S. income statistics since 1948.

The data are publicly available, freely accessible, and licensed for
unrestricted use as a U.S. government work in the public domain. We
accessed the publication on April 30, 2026 from
<https://www.census.gov/library/publications/2025/demo/p60-285.html>.

## FAIR and CARE Assessment

The CPS ASEC data set scores well on each of the FAIR principles for
data stewardship [@wilkinson2016]. The data are *findable* through
persistent Census Bureau publication URLs and report numbers
(P60-285); they are *accessible* without authentication or fees;
they are *interoperable* in standard CSV and Excel formats with rich
column metadata; and they are *reusable* under the public-domain
licensing that applies to all U.S. federal government works.

The CARE principles for Indigenous Data Governance [@carroll2020] are
not directly applicable to this data set because CPS ASEC is a
general-population household survey and does not contain Indigenous
community data subject to Indigenous data sovereignty. We note this
explicitly rather than skip the assessment, because the act of
checking is part of CARE.

## Data Wrangling

The Census Bureau publishes Table A-1 in a format that mixes section
headers with data rows, encodes household and dollar values as
comma-separated text, and embeds statistical-significance flags as
asterisks within otherwise-numeric percent-change columns. Each
section below applies a piped wrangling sequence to its data slice:
rename columns to consistent snake-case names, drop aggregate rows
that would double-count detail rows, capture the asterisk flags into
dedicated logical columns, and use `parse_number()` from `{readr}` to
coerce character-formatted numeric columns into proper numeric type.

{{< pagebreak >}}

# Findings

## Educational Attainment

The education results show a clear earnings premium associated with
higher educational attainment. As shown in @tbl-educationIncome,
households headed by individuals with a bachelor's degree or higher
earned substantially more than all other education groups in both
years, reinforcing the strong income advantage of college completion.
Those with some college also earned more than high school graduates,
though the difference was much smaller.

A notable exception appears among households without a high school
diploma. This group was the only education category to experience a
decline in real median income from 2023 to 2024, indicating that
households with the lowest educational attainment became slightly
worse off in inflation-adjusted terms. In contrast, all other
education groups saw modest gains, suggesting that income growth in
2024 was concentrated among households with at least a high school
education. @fig-educationPremium plots the earnings premium relative
to high school graduates, showing the gap between higher-education
households and the high-school baseline.

```{r}
#| label: dataEducationWrangle

# Primary Author: William Dong
# Reviewer: Bingjie Yan
# ---- Build education income data and earnings premium ----

eduSummary <- data.frame(
  education = c(
    "No high school diploma",
    "High school graduate",
    "Some college",
    "Bachelor's degree or higher"
  ),
  income_2023 = c(37570, 57250, 75510, 130100),
  income_2024 = c(36900, 58410, 76520, 132700),
  percent_change = c(-1.8, 2.0, 1.3, 2.0)
)

# Compute earnings premium relative to high school graduates ----
hs_2023 <- eduSummary$income_2023[2]
hs_2024 <- eduSummary$income_2024[2]

eduPremium <- eduSummary |>
  mutate(
    premium_2023 = (income_2023 / hs_2023) - 1,
    premium_2024 = (income_2024 / hs_2024) - 1,
    premium_delta = premium_2024 - premium_2023
  )
```

```{r}
#| label: tbl-educationIncome
#| tbl-cap: "Median Household Income by Educational Attainment, 2023-2024"
#| tbl-pos: H
#| html-table-processing: none

# Primary Author: William Dong
# Reviewer: Bingjie Yan
# ---- Build education income table ----

eduSummary |>
  mutate(
    pct_change_label = paste0(
      if_else(condition = (percent_change > 0), true = "+", false = ""),
      format(percent_change, nsmall = 1),
      "%"
    )
  ) |>
  select(
    education,
    income_2023,
    income_2024,
    pct_change_label
  ) |>
  kable(
    digits = 0,
    col.names = c(
      "Education Level",
      "Median 2023",
      "Median 2024",
      "Percent Change"
    ),
    format.args = list(big.mark = ","),
    align = "lccc",
    booktabs = TRUE
  ) |>
  kable_classic(latex_options = c("HOLD_position")) |>
  footnote(
    general = paste(
      "Median income reported in 2024 dollars (CPI-U adjusted).",
      "Source: U.S. Census Bureau, CPS ASEC, Table A-1."
    ),
    general_title = "",
    threeparttable = TRUE
  )
```

```{r}
#| label: fig-educationPremium
#| fig-cap: "Higher Education Earnings Premium Relative to High School Graduates, 2023 vs 2024"
#| fig-pos: H
#| fig-alt: "Bar chart showing the earnings premium relative to high school graduates for three education levels in 2023 and 2024. Bachelor's degree or higher shows the largest premium at about 130 percent above high school, followed by some college at about 30 percent. Households with no high school diploma earn about 35 percent less than the baseline."
#| fig-height: 3.8

# Primary Author: William Dong
# Reviewer: Bingjie Yan
# ---- Build education premium plot ----

eduPlot <- eduPremium |>
  filter(education != "High school graduate") |>
  pivot_longer(
    cols = c(premium_2023, premium_2024),
    names_to = "Year",
    values_to = "Premium"
  ) |>
  mutate(
    Year = if_else(Year == "premium_2023", "2023", "2024"),
    education = factor(
      education,
      levels = c(
        "No high school diploma",
        "Some college",
        "Bachelor's degree or higher"
      )
    )
  )

ggplot(
  data = eduPlot,
  mapping = aes(x = education, y = Premium, fill = Year)
) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(values = c("2023" = "#BC204B", "2024" = "#1E407C")) +
  labs(
    x = "Education Level",
    y = "Premium Relative to High School",
    fill = "Year",
    caption = "Source: U.S. Census Bureau, CPS ASEC, Table A-1."
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.caption = element_text(hjust = 0)
  )
```

## Gender of Householder

The gender comparison shows that male-householder households earned
more than female-householder households in both 2023 and 2024, as
reported in @tbl-genderIncome. Although median income remained higher
for men, female-householder households experienced income growth
between the two years while male-householder households experienced a
slight decline.

As a result, the female-to-male income ratio improved in 2024,
indicating that the gender income gap narrowed slightly.
@fig-genderRatio visualizes this comparison directly. Despite this
improvement, the gap remains substantial, with female-householder
households continuing to earn considerably less than male-householder
households overall.

```{r}
#| label: dataGenderWrangle

# Primary Author: Bingjie Yan
# Reviewer: Ethan Tluszcz
# ---- Build gender income data and ratio ----

genderDf <- data.frame(
  group = c("Male Householder", "Female Householder"),
  income_2023 = c(58680, 43230),
  income_2024 = c(58000, 44870),
  moe_2024 = c(1866, 1256)
)

# Calculate female-to-male ratio for each year ----
ratio_2023 <- genderDf$income_2023[2] / genderDf$income_2023[1]
ratio_2024 <- genderDf$income_2024[2] / genderDf$income_2024[1]

genderRatio <- data.frame(
  year = c("2023", "2024"),
  ratio = c(ratio_2023, ratio_2024)
)
```

```{r}
#| label: tbl-genderIncome
#| tbl-cap: "Median Household Income by Gender of Householder, Nonfamily Households, 2023-2024"
#| tbl-pos: H
#| html-table-processing: none

# Primary Author: Bingjie Yan
# Reviewer: Ethan Tluszcz
# ---- Build gender income table ----

genderDf |>
  mutate(
    percent_change = (income_2024 - income_2023) / income_2023 * 100,
    pct_change_label = paste0(
      if_else(condition = (percent_change > 0), true = "+", false = ""),
      format(round(percent_change, 1), nsmall = 1),
      "%"
    )
  ) |>
  select(
    group,
    income_2023,
    income_2024,
    pct_change_label
  ) |>
  kable(
    digits = 0,
    col.names = c(
      "Householder",
      "Median 2023",
      "Median 2024",
      "Percent Change"
    ),
    format.args = list(big.mark = ","),
    align = "lccc",
    booktabs = TRUE
  ) |>
  kable_classic(latex_options = c("HOLD_position")) |>
  footnote(
    general = paste(
      "Median income reported in 2024 dollars (CPI-U adjusted).",
      "Source: U.S. Census Bureau, CPS ASEC, Table A-1."
    ),
    general_title = "",
    threeparttable = TRUE
  )
```

```{r}
#| label: fig-genderRatio
#| fig-cap: "Median Income Comparison: Male vs Female Householders, 2023 vs 2024"
#| fig-pos: H
#| fig-alt: "Grouped bar chart comparing median income for male and female nonfamily householders in 2023 and 2024. Male householders earn about 58,000 dollars in both years; female householders earn about 43,000 in 2023 and about 45,000 in 2024, narrowing the gap slightly."
#| fig-height: 3.8

# Primary Author: Bingjie Yan
# Reviewer: Ethan Tluszcz
# ---- Build gender income comparison plot ----

genderPlot <- genderDf |>
  pivot_longer(
    cols = starts_with("income"),
    names_to = "Year",
    values_to = "Income"
  ) |>
  mutate(
    Year = if_else(Year == "income_2023", "2023", "2024")
  )

ggplot(
  data = genderPlot,
  mapping = aes(x = group, y = Income, fill = Year)
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    mapping = aes(label = label_dollar()(Income)),
    position = position_dodge(0.9),
    vjust = -0.5,
    size = 3.2
  ) +
  scale_y_continuous(
    labels = label_dollar(),
    limits = c(0, 70000)
  ) +
  scale_fill_manual(values = c("2023" = "#BC204B", "2024" = "#1E407C")) +
  labs(
    x = "Householder",
    y = "Median Household Income (2024 dollars)",
    fill = "Year",
    caption = "Source: U.S. Census Bureau, CPS ASEC, Table A-1."
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.caption = element_text(hjust = 0)
  )
```

## Age of Householder

Median household income follows a clear life-cycle pattern across age
groups. As shown in @tbl-ageIncome, income climbs from \$60,310 among
15--24-year-old householders in 2024 to a peak of \$116,800 among
those aged 45--54, before dropping to \$56,680 for households aged 65
and older. @fig-ageIncomeCurve plots both years as overlaid lines so
the year-over-year change at each life stage is easy to read.

```{r}
#| label: dataAgeWrangle

# Primary Author: Ethan Tluszcz
# Reviewer: William Dong
# ---- Read and clean Census age income data ----

# Step 1: Read the raw CSV
ageRaw <- read.csv("age_income_raw.csv")

# Step 2: Wrangle in a piped sequence
ageIncome <- ageRaw |>
  rename(
    age_group = characteristic,
    households_2023 = Population_2023,
    median_2023 = Income_2023,
    moe_2023 = MOE_2023,
    households_2024 = Population_2024,
    median_2024 = Income_2024,
    moe_2024 = MOE_2024,
    pct_change = pct_change_Income,
    pct_change_moe = pct_change_MOE
  ) |>
  filter(age_group != "Under 65 years") |>
  mutate(
    significant_change = if_else(
      condition = (age_group %in% c(
        "15 to 24 years",
        "35 to 44 years",
        "45 to 54 years"
      )),
      true = TRUE,
      false = FALSE
    ),
    pct_change = parse_number(pct_change)
  )
```

```{r}
#| label: tbl-ageIncome
#| tbl-cap: "Median Household Income by Age of Householder, 2023-2024"
#| tbl-pos: H
#| html-table-processing: none

# Primary Author: Ethan Tluszcz
# Reviewer: William Dong
# ---- Build age income table ----

ageIncome |>
  mutate(
    pct_change_label = paste0(
      if_else(condition = (pct_change > 0), true = "+", false = ""),
      format(pct_change, nsmall = 1),
      "%",
      if_else(condition = significant_change, true = "*", false = "")
    )
  ) |>
  select(
    age_group,
    households_2023,
    median_2023,
    households_2024,
    median_2024,
    pct_change_label
  ) |>
  kable(
    digits = 0,
    col.names = c(
      "Age",
      "Households 2023",
      "Median 2023",
      "Households 2024",
      "Median 2024",
      "Percent Change"
    ),
    format.args = list(big.mark = ","),
    align = "lccccc",
    booktabs = TRUE
  ) |>
  kable_classic(latex_options = c("HOLD_position", "scale_down")) |>
  footnote(
    general = paste(
      "Household counts are reported in thousands.",
      "Median income is reported in 2024 dollars (CPI-U adjusted).",
      "Asterisks indicate that the year-over-year change in real income",
      "was statistically significant at the 90% confidence level.",
      "Source: U.S. Census Bureau, CPS ASEC, Table A-1."
    ),
    general_title = "",
    threeparttable = TRUE
  )
```

```{r}
#| label: fig-ageIncomeCurve
#| fig-cap: "Median Household Income by Age of Householder, 2023 vs 2024"
#| fig-pos: H
#| fig-alt: "Line graph showing median U.S. household income by age of householder for 2023 and 2024. Both years follow an inverted-U curve, peaking among households aged 45 to 54 around $116,000 and falling sharply at retirement. The 15-24 group gained the most year-over-year, +7.0%, statistically significant."
#| fig-height: 4.5

# Primary Author: Ethan Tluszcz
# Reviewer: William Dong
# ---- Build age income line chart ----

# Reshape to long format so year can map to color ----
ageIncomeLong <- ageIncome |>
  pivot_longer(
    cols = c(median_2023, median_2024),
    names_to = "year",
    values_to = "median_income"
  ) |>
  mutate(
    year = if_else(
      condition = (year == "median_2023"),
      true = "2023",
      false = "2024"
    )
  )

# Set the order of age groups for the x-axis ----
ageOrder <- c(
  "15 to 24 years",
  "25 to 34 years",
  "35 to 44 years",
  "45 to 54 years",
  "55 to 64 years",
  "65 years and older"
)

# Short labels for the x-axis (avoids overlap) ----
ageLabels <- c(
  "15 to 24 years"     = "15-24",
  "25 to 34 years"     = "25-34",
  "35 to 44 years"     = "35-44",
  "45 to 54 years"     = "45-54",
  "55 to 64 years"     = "55-64",
  "65 years and older" = "65+"
)

# Build the plot ----
ggplot(
  data = ageIncomeLong,
  mapping = aes(
    x = age_group,
    y = median_income,
    color = year,
    group = year
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_text(
    data = ageIncome,
    mapping = aes(
      x = age_group,
      y = if_else(
        condition = (median_2024 > median_2023),
        true = median_2024,
        false = median_2023
      ),
      label = paste0(
        if_else(condition = (pct_change > 0), true = "+", false = ""),
        format(pct_change, nsmall = 1),
        "%",
        if_else(condition = significant_change, true = "*", false = "")
      )
    ),
    inherit.aes = FALSE,
    vjust = -1.2,
    size = 3.2,
    fontface = "bold"
  ) +
  scale_x_discrete(limits = ageOrder, labels = ageLabels) +
  scale_y_continuous(
    labels = label_dollar(),
    limits = c(40000, 130000)
  ) +
  scale_color_manual(
    values = c("2023" = "#BC204B", "2024" = "#1E407C")
  ) +
  labs(
    x = "Age of Householder",
    y = "Median Household Income (2024 dollars)",
    color = "Year",
    caption = paste0(
      "Source: U.S. Census Bureau, CPS ASEC, Table A-1.\n",
      "Labels show real percent change in median income, 2024 less 2023; ",
      "* indicates statistically significant change."
    )
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.caption = element_text(hjust = 0)
  )
```

The most striking feature of @fig-ageIncomeCurve is the inverted-U
shape of the income-age curve, which is consistent across both years.
Median income roughly doubles between the youngest and peak-earning
groups (\$60,310 versus \$116,800 in 2024), and then drops by half
again at retirement.

Year-over-year growth was strongest among the youngest households,
where the 15--24 group gained 7.0% in real terms — the largest gain
in the table, and statistically significant at the 90% confidence
level — suggesting that the labor-market gains of 2024 were
concentrated among younger workers entering the labor market. The
35--44 and 45--54 groups also posted significant real-income gains of
2.1% and 2.8% respectively. Only the 55--64 group lost ground in real
terms, declining 1.5%, though that change was not flagged as
statistically significant by the Census Bureau and so should be
interpreted as noise rather than a real reversal.

For drafting the alt text and long description for @fig-ageIncomeCurve,
we used ASU's Image Accessibility AI Tool [@asuImageAccessibility].
The tool generated a candidate long description from the rendered
chart, which we then reviewed and corrected before incorporation.
Specifically, the tool initially attributed the line colors backwards
(mapping red to 2024 and blue to 2023, when the chart actually uses
red for 2023 and blue for 2024), and we edited the description
manually to fix this and to add the statistical significance
asterisks the tool omitted.

{{< pagebreak >}}

# Discussion

Across the three demographic characteristics examined in this report,
the most consistent pattern is that the largest income disparities
remain structural rather than driven by year-over-year change. The
gap between bachelor's-degree-or-higher households and households
without a high school diploma is several times larger than the gap
between any two consecutive age groups, and the gap between male and
female nonfamily householders, while smaller in absolute terms, has
narrowed more in percentage terms than either of the other two
disparities. Real-income changes between 2023 and 2024 were modest at
the population level, but the gains were unevenly distributed:
younger householders, households with a bachelor's degree or higher,
and female nonfamily householders all saw real gains, while the
55--64 age band and households without a high school diploma were
the only groups examined here to lose ground in real terms.

We are appropriately cautious about extrapolating from a two-year
window. The CPS ASEC publishes one observation per year per
characteristic, so this analysis captures a short-term snapshot
rather than a long-term structural trend. A longer time series would
be needed to distinguish a one-year fluctuation from a sustained
shift, particularly given how recent decades have been shaped by
major economic shocks: the dot-com bust, the 2008 financial crisis,
and the COVID-19 pandemic each reshaped both the labor market and
the income distribution. Future work could extend this analysis to a
multi-year window and incorporate unemployment rates as a covariate
to test whether age-, gender-, and education-specific income changes
correlate with broader labor-market conditions.

# Author Contributions

This project was a collaborative effort by all three authors,
following the CRediT contributor taxonomy [@brand2015]:

- **William Dong:** Conceptualization (lead); Data Curation
  (educational attainment); Investigation (educational attainment);
  Visualization (Table 1, Figure 1); Writing -- Original Draft
  (educational attainment section).
- **Bingjie Yan:** Conceptualization (supporting); Data Curation
  (gender); Investigation (gender); Visualization (Table 2,
  Figure 2); Writing -- Original Draft (gender section).
- **Ethan Tluszcz:** Data Curation (age of householder);
  Investigation (age of householder); Methodology (data wrangling
  pipeline, code style); Visualization (Table 3, Figure 3); Writing
  -- Original Draft (introduction, data and methods, discussion).

All authors reviewed code via Pull Request, contributed to the
discussion section, and approved the final manuscript before
submission.

{{< pagebreak >}}

# Appendix — Code

```{r}
#| label: codeAppend
#| ref-label: !expr knitr::all_labels()
#| echo: true
#| eval: false
```
